Handwriting Recognition

### 2017-12-17 Report/paper.tex

\vskip 0.3in
]
% Notes:
% - Add https://tel.archives-ouvertes.fr/tel-01249405/document reference, (see abstract) 
% in order to compare results. 
% - Disciss possible flaws. (Use images from IAM in order to do so)
% - Update outdated parts, like neural net of splitpoint evaluation
%  Reference datasets, Chars74K, IAM, Google n-grams
%  redo the preprocessing experiment and explain why some techniques increase the accuracy more than others. 
\begin{abstract} 
The aim of this paper is to present a system for off-line handwriting recognition. 
We discuss how to segment texts with contouring. Next we present our method of splitting words into characters. 
Then we describe how to recognise characters with a deep neural network approach. 
Finally we present some postprocessing techniques to correct mispredicted text. 
\end{abstract}

@@

First we explain how to recognise characters in Section \ref{sec:charrec}. This explanation involves the used dataset, preprocessing techniques and the deep neural network for classifying characters. 
In Section \ref{sec:segtext} we describe how to segment a text into words using a contouring algorithm with consequent postprocessing to remove unwanted noise. 
After this we present our approach to seperating the characters from words in \ref{seg:segword}. We use vertical projections combined with postprocessing techniques including a neural network for splitpoint evaluation.
Postprocessing steps are presented in Section \ref{seg:postproc}, we try to correct word recognition errors using an English dictionary and language models, n-gram models in particular.
We conclude this paper with some experimental results, the problems we encountered combined with possible improvements.

@@

\section{Segmentation of text into words}
\ref{sec:segtext}
\section{Segmentation of words into characters}
\ref{sec:segword}
\subsection{Correction of oversegmentation}
As our algorithm for segmentation of handwritten words splits the word into to many characters, we can use techniques to verify the correctness of these splitting points. 
\cite{evalsplitpoints} discusses three approaches to evaluate split point correctness: rule based techniques, cost function optimization based methods and machine learning based procedures. 
In this paper we will only discuss the latter, machine learning based procedures, as we used this approach. 
In particular, our algorithm is based on \cite{evalsplitpointsnn}. The paper explains how to use neural networks to classify the correctness of splitting points. The discussed algorithm is as follows. Manual classification of splitting points creates training data with two classes, correct and incorrect splitting points. 
For every of these splitting points a pixel matrix is extracted surrounding the splitting point. This pixel matrix is normalised in size. 
Next, small windows of equal size are extracted from the pixel matrix and for each window the density is calculated. This is the number of black pixels divided by the total amount of pixels in the window. 
With every matrix a corresponding label is encoded for storage in the training file, this label is encoded as 0.9 if the splitting point is correct and 0.1 otherwise. Now these matrices and corresponding labels are used to train the neural network.

We take a slightly different approach and consider a deep neural network to tackle this problem. This allows us to use convolutational layers in order to extract features, instead of manually creating features (features which show the density in this case). 
We use one convolutional layer and three fully connected layers in our neural network. This neural network takes the actual image of the splitting point as input (a pixel matrix of the pixels surrounding the splitting point) and returns a one-hot encoded vector indicating if the splitting is correct or incorrect. 
This method allowed us to achieve a higher accuracy than an implementation of the method described by \cite{evalsplitpointsnn}.
Splitting points can now be filtered if the neural network decides a splitting point is incorrect.

>>>>

\vskip 0.3in
]
% Notes:
% - Add https://tel.archives-ouvertes.fr/tel-01249405/document reference, (see abstract) 
% in order to compare results. 
% - Disciss possible flaws. (Use images from IAM in order to do so)
% - Update outdated parts, like neural net of splitpoint evaluation
%  Reference datasets, Chars74K, IAM, Google n-grams
%  redo the preprocessing experiment and explain why some techniques increase the accuracy more than others. 
\begin{abstract} 
The aim of this paper is to present a system for off-line handwriting recognition. 
We discuss how to segment texts with contouring. Next we present our method of segmentating words into characters. 
Then we describe how to recognise characters with a deep neural network approach. 
Finally we present some postprocessing techniques to correct mispredicted text. 
\end{abstract}

@@

First we explain how to recognise characters in Section \ref{sec:charrec}. This explanation involves the used dataset, preprocessing techniques and the deep neural network for classifying characters. 
In Section \ref{sec:segtext} we describe how to segment a text into words using a contouring algorithm with consequent postprocessing to remove unwanted noise. 
After this we present our approach to seperating the characters from words in \ref{seg:segword}. We use vertical projections combined with postprocessing techniques including a neural network for segmentation point evaluation.
Postprocessing steps are presented in Section \ref{seg:postproc}, we try to correct word recognition errors using an English dictionary and language models, n-gram models in particular.
We conclude this paper with some experimental results, the problems we encountered combined with possible improvements.

@@

\section{Segmentation of text into words}
\ref{sec:segtext}

To be able to recognise the characters in a given text, we will first need to extract the words from the text before we can start evaluating the characters. This word extraction has many possible approaches, but in this paper we will focus solely on the word contouring approach.
This method uses a heuristic algorithm to extract the contours of the words from a given text, and through postprocessing we can rearrange these extracted words into lines.
The algorithm proposed is only to processes text images, and does not take into account images and other non-textual areas found in the text.
Dependant on the situation, additional techniques for page segmentation, such as the Run-Length Smearing Algorithm, can be utilized.
% TODO:: Run length Smearing Algorithm reference --> ....

\subsection{binarization}
The first preprocessing step is binarizing the text image. 
If the image is not yet greyscaled, this is done at the start of the preprocessing phase.
The greyscaled image is converted from greyscaled into a binary image.
This process will allows us to easily separate the text from the background.

To achieve this there are a lot of thresholding algorithms to choose from, but the one we will use throughout the paper is a combination of executing Gaussian blur in combination with OTSU's thresholding algorithm on the binary text image.
The algorithms used are as implemented in the python opencv library. %TODO:: where do i put this????

The Gaussian blur will reduce the noise in the image, and smoothen the edges of the text.
The smoothening of the background noise will result in less noise picked up by the chosen thresholding algorithm, so it is less likely to confuse background noise with the words from the text image.

The tresholding algorithm used in this paper is OTSU's thresholding algorithm.
This algorithm will deliver adequate result for noisy text documents.
The choise of tresholding algorithm depends largely on intended purposes of the algorithm.
If very noisy documents will need to be processed, additional noise removal techniques and other preprocessing techniques may be required to reduce the excess noise picked up by the tresholding algorithm.

\subsection{contouring}
Our heuristic text segmentation algorithm is based on finding the contours in the text image, so we can group these contours in postprocessing stages to recover the original words from the text image.
The contouring algorithm used is the one implemented in the opencv library for python.
This algorithm generates a lot of unneeded contours, which we will filter through postprocessing.
Also since words in handwritten text are not always connected into one piece, we will need to connect these contours in the postprocessing stages as well.

\subsection{postprocessing}
In the postprocessing step, we eliminate the redundant extracted contours, and reconnect the contours of words that consist of multiple parts.

\subsubsection{contour removal}
Since the contouring algorithm produces contours that are not contours of the words in the text, these contours need to be filtered out and removed.
To accomplish this we make use of the bounding box we can draw around a contour.
This bounding box is the smallest rectangle containing said contour, and it will allow us to easily compare contour positions in the text image.

The first filter is removing the contours of which the bounding box is contained by the bounding box of another contour.
This is allowed because the contents of this inner bounding box will be regarded when processing the contents of the outer bounding box, and so we can remove the contours belonging to the inner bounding boxes.

The second filter is removing the contours that are too small to contain parts of a word.
Since the contouring algorithm extracts the contours for any size, and because our character recognition network does not process punctuation, 
we need to eliminate these from our results.

We achieve this by taking the average height of all the found contours in the text.
From these contours we calculate the average height, after which we remove the contours with a height lesser than the average height multiplied by a chosen parameter $minimalHeightMultiplier$.
The optimal results were achieved when we initialize with $minimalHeightMultiplier = 2/3$, so that we retain only contours higher than the average height $ * 2/3$.

\subsubsection{contour combination}
With the excess contours removed, we can begin to reconstruct the original words from the text by combining multiple contours into one where needed.
The algorithm used for this is a straightforward overlapping algorithm, that combines word segments based on the overlap between the bounding box of their contours.
The parameters for this algorithm are the $alowedWordGap$ and the $allowedVerticalOverlap$ paramters.
These parameters are initialized as $interWordSegmentDistance = $ the average width of the found words $ / 16 $ and $allowedVerticalOverlap = 1/2$ to give the best results in the our testset, however further experimentation with these parameters and the algorithm is required to find a more generalized optimal solution.

\subsection{line selection}
Now that the words are extracted from the text image, the next goal is to try and reconstruct the original text by reordering these words in their original lines.
For the line selection process, an iterative algorithm, Algorithm \ref{alg:linereconstr}, is used.
\begin{algorithm}[tb]
   \caption{Line reconstruction algorithm}
   \label{alg:linereconstr}
\begin{algorithmic}
   \STATE {\bfseries Input:} $extractedWords$, $textlines$
   \REPEAT
        \STATE $highestWord = findHighestWord $
        \STATE $newline = [highestWord]$
        \STATE Initialize $noChange = true$.
        \STATE $avgY= newline$.findAverageYpos
        \REPEAT
            \FOR{$word$ {\bfseries in} $extractedWords$}
            \STATE $boundRect = word.contour.boundingRect$
            \IF{ $avgY > word.y$ {\bfseries and} $avgY < word.y + word.h$ } 
            \STATE $line$.add $word$
            \STATE $noChange = false$
            \ENDIF
        \ENDFOR
        \UNTIL{$noChange = true$}
        \STATE $textlines$.add $newline$
   \UNTIL{$words$.empty $= true$}
\end{algorithmic}
\end{algorithm}

This algorithm reorders the found words in lines, and return these lines as ordered in the original text image.

This algorithm fares well processing text images, where the text consists of parrallel text lines.
Angeled parallel text lines can be compensated for by rotating and padding the thresholded text image with different angles, and comparing the results.

Combinations of parrallel and slanted text lines however are not handled well by the algorithm, and will require further adaptations or more advanced algorithms to map the words to text lines correctly.

\section{Segmentation of words into characters}
\ref{sec:segword}
The next step in the process is to extract the character images from the given word images.
Again, there is the choise between several different approaches.
In this paper we focus solely on a word oversegmentation approach to character extraction.
The purpose is to segment the words into the individual characters.
This will be done through initial oversegmentation of the word image, followed by postprocessing steps to remove the excess separation points to recover the original character images.

\subsection{Normalization} % TODO:: necessary? never taken advantage of? maybe for split points count etc???
The first action is normalizing the word image.
This normalization step enables us to reason about words written in different pixel scalings the same way.
The process of normalizing the word image consists of resizing the image to a fixed height of $100px$.

\subsection{Binarization}
Next we will again need to execute a binarization algorithm on the word image.
Before thresholding the image, Gaussian blur is executed to reduce the noise of the word image, and to smoothen the edges of the characters.
Afterwards we use OTSU's thresholding algorithm again to binarize the blurred word image.
This thresholding step is essential for out skeletonization algorithm in the next step, as it requires a binarized image as input.

\subsection{Skeletonization}
The skeletonization of the thresholded word image will allow us to ignore the thickness of the lines in further processing. 
This normalization is needed to reason about potential segmentation points between characters in the word image.
After comparing different skeletonization algorithms, the algorithm of choise is the skeletonize function found in the scipy python library.
This is an adaptation of the Zhang-Suen thinning algorithm, one of the most commonly used and adapted thinning algorithms.
%TODO:: GIVE REFERENCE TO ZHANG SUEN AND SCIPY MANUAL FOR SKELETONIZATION

\subsection{Vertical projection histogram}
Now that we have the skeletonized word image, we can proceed with determining the potential segmentation points.
We do this by taking the vertical projection histogram of this skeletonized word image.
Before calculating the projection histogram, we first have to normalize the image to contain values $1$ for black pixels and $0$ for white pixels in the image.
When the projection vertical projection is executed, we have the summation of every collumn of the word image.
Because the word image is skeletonized, we can regard the collumns with a total pixel count of zero or one pixel as potential segmentation points, because connections between handwritten words consist of a single stroke, or whiitespace if the word consists of multiple parts.

Now that we have our potential segmentation points, we need to decide which ones are correct segmentation points, and which ones need to be removed. This is done through correction of our oversegmentation.

\subsection{Oversegmentation correction}
In this section we provide a method to fix the oversegmentation of the word caused by our previous steps and to decide which segmentation points to retain.
To achieve this we first process the potential segmentation points, to remove redundant and incorrect segmentation points.
THe final decision to which segmentation points we retain is made by a neural network.

\subsubsection{rule based}
The main oversegmentation correction is handled by our neural network.
However some simple solutions can be added to remove unwanted potential segmentation points throught deterministic processing.

The first filter we utilize here, is one that removes the segmentation points at the start and the end of the word image.
These segmentation points are an artifact of the used segmentation approach, and can be removed without information loss, as the alphabet does not contain letters that convert to a single pixel line after skeletonization.

A second filter groups consequent potential segmentation points.
The occurence of these consequent points is another artifact from our segmentation approach.
From these grouped points, the collumns beloning to these x-coordinates are checked to be empty or containing a single pixel in the vertical projection histogram.
if the group contains collumns that are empty on the histogram, only the middle segmentation point from this group is retained as potential segmentation point.
If none of the potential segmentation points has an empty collumn on the histogram, the retained segmentation point is the middle points from the group.

A third filter is that groups of consequent potential segmentation points, separated by collumns containing less pixels than a certain threshold value $maxSumThresh$, are grouped together.
The value chosen for this threshold is 3.
This filter remedies the segmentation of slated lines between the characters.

This value for this threshold is pretty conservative, and could be further investigated for a more optimal rule based approach.
More rule based filter could be added for more optimal processing of the potential segmentation points.
However the focus the segmentation correction is the neural network used to remove incorrect segmenatation points.

%As our algorithm for segmentation of handwritten words splits the word into to many characters, we can use techniques to verify the correctness of these segmentation points points. 
%\cite{evalsplitpoints} discusses three approaches to evaluate segmentation point correctness: rule based techniques, cost function optimization based methods and machine learning based procedures. 
%In this paper we will only discuss the latter, machine learning based procedures, as we used this approach.

\subsubsection{Neural network}
The neural network we use to judge the potential segmentation points, is based on \cite{evalsplitpointsnn}. The paper explains how to use neural networks to classify the correctness of segmentation points. The discussed algorithm is as follows. Manual classification of segmentation points creates training data with two classes, correct and incorrect segmentation points. 
For every of these segmentation points a pixel matrix is extracted surrounding the segmentation point. This pixel matrix is normalised in size. 
Next, small windows of equal size are extracted from the pixel matrix and for each window the density is calculated. This is the number of black pixels divided by the total amount of pixels in the window. 
With every matrix a corresponding label is encoded for storage in the training file, this label is encoded as 0.9 if the segmentation point is correct and 0.1 otherwise. Now these matrices and corresponding labels are used to train the neural network.

We take a slightly different approach and consider a deep neural network to tackle this problem. This allows us to use convolutational layers in order to extract features, instead of manually creating features (features which show the density in this case). 
We use one convolutional layer and three fully connected layers in our neural network. This neural network takes the actual image of the segmentation point as input (a pixel matrix of the pixels surrounding the segmentation point) and returns a one-hot encoded vector indicating if the segmentation is correct or incorrect. 
This method allowed us to achieve a higher accuracy than an implementation of the method described by \cite{evalsplitpointsnn}.
Segmentation points can now be filtered if the neural network decides a segmentation point is incorrect.

\subsection{Normalization}
The character normalization step consists of two parts.
Firstly the character is padded untill the height equals the width.
A precondition for the padding step is that the image is already binarized, so that we can use white borders as padding, and this will not distort the image should binarization take place in a later stage.
The image is then resized to $44px x 44px$
Secondly padding of $10px$ is added on all sides of the character image to create whitespace around the letter.
We end up with a $64px x 64px$ character image.
The extra $10px$ padding is to make sure the letters do not stick to the edges, since the Chars74K dataset we work with utilizes borders around the provided characters. Should these borders be removed in the training process however, this should be reflected in this step by ignoring this new padding.
